---
title: "Statistical inference with the GSS data"
output: rmarkdown::github_document
#output: 
#  html_document: 
#    fig_height: 4
#    highlight: pygments
#    theme: spacelab
---

## Setup

### Load packages

```{r load-packages, message = FALSE}
library(ggplot2)
library(dplyr)
library(statsr)
```

### Load data

Make sure your data and R Markdown files are in the same directory. When loaded
your data file will be called `gss`. Delete this note when before you submit 
your work. 

```{r load-data}
load("gss.Rdata")
```

* * *

## Part 1: Data

### Background
This project uses an extract of the General Social Survey (GSS) Cumulative File 1972-2012.
<br/><br/>
The General Social Survey (GSS) is a nationally representative survey of adults in the United States conducted since 1972. The GSS collects data on contemporary American society in order to monitor and explain trends in opinions, attitudes and behaviors. The GSS has adopted questions from earlier surveys which allows researchers to conduct comparisons for up to 80 years.
<br/>
The GSS contains a standard core of demographic, behavioral, and attitudinal questions, plus topics of special interest. Among the topics covered are civil liberties, crime and violence, intergroup tolerance, morality, national spending priorities, psychological well-being, social mobility, and stress and traumatic events.

### History
The GSS was first conducted in 1972. Until 1994, it was conducted almost annually (with the exceptions of the years 1979, 1981, and 1992). Since 1994, the GSS has been conducted in even numbered years.
In 2006, a large part of the GSS was administered in Spanish for the first time.

### Methodology
The target population of the GSS is adults (18+) living in households in the United States. The GSS sample is drawn using an area probability design that randomly selects respondents in households across the United States to take part in the survey. Respondents that become part of the GSS sample are from a mix of urban, suburban, and rural geographic areas. Participation in the study is strictly voluntary.<br/>
The survey is conducted face-to-face with an in-person interview by NORC at the University of Chicago.

### Scope inference
```{r}
dim(gss)
```

The study is conducted by random sampling with a good size which makes the data generalizable to the US adult population.<br/>
The GSS is an observational study with no random assignments that can have an indication of association among the relationship of the variables of interest, but not causation.<br/><br/>
The GSS is declared unbiased but it may have a set of constraints that introduce bias as:
<ul>
  <li>Introduce Spanish speakers only in 2006</li>
  <li>To be a voluntary survey, making it impossible to report those that decide not to answer it</li>
</ul>

* * *

## Part 2: Research question

Is there a relationship between social class and the belief about black discrimination? Is this belief changing across the years?
How different social classes see their expenses to improve the conditions of black people?

After the crimes committed against some black people in 2020 and the big number of riots against them, using some data is possible to analyze the opinion about structural racism in the US. The study will compare the opinion about discrimination against black people per social class and how that is changed in the last years. It is possible to analyze per social classes people opinion about their expenses to improve the conditions of black people in the country.

* * *

## Part 3: Exploratory data analysis

First, to decrease the number of columns in the sample, it will keep only the variables that can give some insight into the study and remove NaNs.
```{r}
# natrace => EXPENCES TO IMPROVING THE CONDITIONS OF BLACKS

# racdif1 => On the average (negroes/blacks/African-Americans) have worse jobs, income, and housing than white people.
# Do you think these differences are: a. Mainly due to discrimination?
columns <- c('year', 'natrace', 'racdif1', 'class')
race <- na.omit(gss[columns])

head(race, 10)
```

To analyze the data distribution:
```{r}
table(race$natrace, useNA = 'ifany')
table(race$racdif1, useNA = 'ifany')
table(race$class, useNA = 'ifany')
```

The tables show that majority does not believe in race discrimination and believe that their expense is `About Right` to improve the conditions of black people. The majority consider themselves `Middle Class` or `Working Class`.

Comparing the social classes and the belief about race discrimination:
```{r, fig.width=10,fig.height=6}
df <- round(prop.table(table(race$class, race$racdif1), margin = 1) * 100, 2)
df

barplot(df, beside = TRUE, legend = rownames(df), args.legend = list(x = "topright", bty = "n", inset=c(0, 0)), col = rainbow(nrow(df)))
```

Comparing the social classes and the belief about their expenses into black people's conditions:
```{r, fig.width=10,fig.height=6}
df <- round(prop.table(table(race$class, race$natrace), margin = 1) * 100, 2)
df

barplot(df, beside = TRUE, legend = rownames(df), args.legend = list(x = "topright", bty = "n", inset=c(0, 0)), col = rainbow(nrow(df)))
```

Looking at the data is possible to notice as higher as the social class less belief in race discrimination and more acceptance with the current expenses to improve black people's conditions.

```{r message = FALSE, warning = FALSE}
df <- race
df$Yes <- 0
df[df$racdif1 == 'Yes',]$Yes <- 1
df$No <- 0
df[df$racdif1 == 'No',]$No <- 1

df <- df %>% group_by(year, class) %>%  summarise(Yes = sum(Yes), No = sum(No))
df$Yes_Perc <- round(df$Yes / (df$Yes + df$No) * 100, 2)
df$No_Perc <- round(df$No / (df$Yes + df$No) * 100, 2)

lower <- df %>% filter(class == 'Lower Class') %>% select(year, Yes_Perc)
barplot(height = lower$Yes_Perc, names = lower$year,
        col = palette.colors(nrow(lower)),
        main = "Lower Class", xlab="Year", ylab = 'Percentage', las = 2)
abline(h = mean(lower$Yes_Perc), col = 'red', lwd = 2)

working <- df %>% filter(class == 'Working Class') %>% select(year, Yes_Perc)
barplot(height = working$Yes_Perc, names = working$year,
        col = palette.colors(nrow(working)),
        main = "Working Class", xlab="Year", ylab = 'Percentage', las = 2)
abline(h = mean(working$Yes_Perc), col = 'red', lwd = 2)

middle <- df %>% filter(class == 'Middle Class') %>% select(year, Yes_Perc)
barplot(height = middle$Yes_Perc, names = middle$year,
        col = palette.colors(nrow(middle)),
        main = "Middle Class", xlab="Year", ylab = 'Percentage', las = 2)
abline(h = mean(middle$Yes_Perc), col = 'red', lwd = 2)

upper <- df %>% filter(class == 'Upper Class') %>% select(year, Yes_Perc)
barplot(height = upper$Yes_Perc, names = upper$year,
        col = palette.colors(nrow(upper)),
        main = "Upper Class", xlab="Year", ylab = 'Percentage', las = 2)
abline(h = mean(upper$Yes_Perc), col = 'red', lwd = 2)

df %>%
  ggplot( aes(x=year, y=Yes_Perc, group=class, color=class)) +
    geom_line()

df[df$year == 2012,]
```

Observing all data per year is possible to see that the belief of race discrimination is decreasing during the last years.

* * *

## Part 4: Inference

<span style="font-size: 12px; font-weight: bold;">This section executes a statistical inference via hypothesis testing following the steps below:</span>
<ul>
  <li>State hypotheses</li>
  <li>Check conditions</li>
  <li>Methodology</li>
  <li>Perform inference</li>
  <li>Interpret results</li>
</ul>

### Is there a relationship between social class and the belief about black discrimination?

#### State hyphoteses

<span style="font-weight: bold;">Null Hypothesis (Ho): </span> The belief of discrimination against black people and social class are independent.<br/>
<span style="font-weight: bold;">Alternative Hypothesis (Ha): </span> The belief of discrimination against black people and social class are dependent.<br/>

#### Check conditions

<span style="font-weight: bold;">Independence: </span> The GSS dataset is generated from a random sample survey with a size below 10% of the population. Because of that is possible to assume independence in this data sample.<br/>
<span style="font-weight: bold;">Expect count: </span> There are at least 5 elements per cell except in `No Class` which will be removed from this study.<br/>

```{r}
table(race$class, race$racdif1)
```

Remove `No Class`
```{r}
df <- race[race$class != 'No Class',]
df <- droplevels(df)
table(df$class, df$racdif1)
```

#### Methodology
The dataset consists of two categorical variables `class` (social class) and `racdif1` (belief of race discrimination), with a categorical variable with more than 2 levels and another with 2 levels. Because of those characteristics, the method adopted is a chi-square test of independence. This is the method adopted when comparing 2 categorical variables and at least one has more than 2 levels.

#### Perform inference

```{r}
chisq.test(df$class, df$racdif1)
```

#### Interpret results
With a p-value of almost zero, there is have strong evidence to reject the null hypothesis. Then is possible to conclude that exists a dependency between social class and the belief of race discrimination. With that is possible to assume an indication of association among the relationship of the variables of interest, but not causation.

<br/>

### How different social classes see their expenses to improve the conditions of black people?

#### State hyphoteses

<span style="font-weight: bold;">Null Hypothesis (Ho): </span> The expenses to improve the conditions of black people and social class are independent.<br/>
<span style="font-weight: bold;">Alternative Hypothesis (Ha): </span> The expenses to improve the conditions of black people and social class are dependent.<br/>

#### Check conditions

<span style="font-weight: bold;">Independence: </span> The GSS dataset is generated from a random sample survey with a size below 10% of the population. Because of that is possible to assume independence in this data sample.<br/>
<span style="font-weight: bold;">Expect count: </span> There are at least 5 elements per cell except in `No Class` which will be removed from this study.<br/>

```{r}
table(race$class, race$natrace)
```

Remove `No Class`
```{r}
df <- race[race$class != 'No Class',]
df <- droplevels(df)
table(df$class, df$natrace)
```

#### Methodology
The dataset consists of two categorical variables `class` (social class) and `natrace` (expenses to improve the conditions of black people), with both categorical variables with more than 2 levels. Because of those characteristics, the method adopted is a chi-square test of independence. This is the method adopted when comparing 2 categorical variables and at least one has more than 2 levels.

#### Perform inference

```{r}
chisq.test(df$class, df$natrace)
```

#### Interpret results
With a p-value of almost zero, there is have strong evidence to reject the null hypothesis. Then is possible to conclude that exists a dependency between social class and their expenses to improve the conditions of black people. With that is possible to assume an indication of association among the relationship of the variables of interest, but not causation.

<br/>